---
title: "Reading and Using Data in Textual Dataset File Format in R"
package: tidyfr
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Reading and Using Data in Textual Dataset File Format in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

**Authors**: `r packageDescription("tidyfr")[["Author"]] `<br />
**Last modified:** `r file.info("tidyfr.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Introduction

The Textual Dataset File Format (TDFF) is a file format to store data from
biomedical experiments in simple text files which enables easier extension,
storage and editing. A detailed description of the format is provided
[here](https://wiki.gm.eurac.edu/index.php?title=Textual_Dataset_Format). The
[tidyfr](https://github.com/EuracBiomedicalResearch/tidyfr) is an R package
designed to facilitate import and export of data in TDF format to and from R.

## General usage

Data from different *modules* (i.e. data sets/sources) in TDFF are stored in
separate folders under a base data directory. Note that also different versions
for the same data set (module) are stored in separate folders hence allowing to
load data from a specific version of a module.

Below we use some (completely randomized) test data bundled within this package
to illustrate the use of the `tidyfr` package. First we define the base data
folder that contains the data sets.

```{r, message = FALSE}
library(tidyfr)

data_path <- system.file("txt", package = "tidyfr")
```

For *real* use of the package this `data_path` variable should point to the data
directory containing the *real* data. We can next list available modules using
the `list_data_modules` function:

```{r, warning = FALSE}
list_data_modules(data_path)
```

We can now choose one of the modules and load it by specifying its name, version
and the base data path.

```{r, warning = FALSE}
mdl <- data_module(name = "db_example2", version = "1.0.1", path = data_path)
```

Various information from a module can be retrieved using the `moduleName`,
`moduleDate`, `moduleDescription` and `moduleVersion` functions.

```{r}
moduleName(mdl)
moduleDate(mdl)
moduleDescription(mdl)
moduleVersion(mdl)
```

The actual data from a data module can be retrieved with the `data`
function. This function returns a `data.frame` and ensures that the data is
correctly formatted (i.e. categorical variables are represented as factors, with
the right categories, dates are formatted correctly and missing values are
represented by `NA`). The AIDs (identifiers of the samples) provided by the data
module are used as row names of the `data.frame` (if these AIDs are not unique,
i.e., some AIDs are duplicated, AIDs are **not** used as row names but provided
in an additional column `"aid"` in the returned `data.frame`).

```{r}
d <- data(mdl)
d
```

Categorical variables are correctly converted to factors:

```{r}
d$x0_sex
```

Dates are converted into the correct format:

```{r}
d$x0_examd
```

Annotations (labels) for the individual variables can be retrieved using the
`labels` function:

```{r}
labels(mdl)
```

The `groups` function retrieves the optional grouping of variables and the
`grp_labels` function the description/definition of these groups.

```{r}
groups(mdl)
grp_labels(mdl)
```


## Data export

Data can be exported using the `export_tdf` function. See the help for this
function for more details.

# Session information

```{r}
sessionInfo()
```
