---
title: "Using the CHRIS p180 targeted metabolomics TDFF(ormat) data in R"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
---

```{r  biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Introduction

This document describes how the Biocrates p180-based targeted metabolomics data
from the CHRIS study [@pattaro_cooperative_2015] [@verri_hernandes_age_2022] can
be loaded and analyzed in R.

The provided *values* for each metabolite is a absolute concentration in
**natural scale**.


# Usage

Data in TDF format is best imported into R using the
[tidyfr](https://github.com/EuracBiomedicalResearch/tidyfr) R package which can
be installed using the code below.

```{r, eval = FALSE}
remotes::install_github("EuracBiomedicalResearch/tidyfr")
```

To load the CHRIS p180 metabolomics data, access to the folder with the data in
TDF format is required. The name of the data folder (and hence the name of the
*CHRIS module*) is *metabolomics_p180*. Below we list all versions of the data
set available for the this module. In this example we assume the
*metabolomics_p180* folder to be present in the current folder in which R is
run. Parameter `path` allows to specify the folder in which CHRIS data modules
are stored (we use `"."` to indicate the current working directory).

```{r}
library(tidyfr)

list_data_modules(path = ".")
```

In this particular case, 2 versions of the module are available. We can next
load the module with the `data_module` function specifying the name of the
module, the version we want to load and also the path where the module can be
found.

```{r}
metabo <- data_module("metabolomics_p180", version = "1.0.1.1", path = ".")
```

The `metabo` variable is now a reference to this module which provides some
general information:

```{r}
metabo
```

Note that we could also use the functions `moduleName`, `moduleVersion`,
`moduleDescription` and `moduleDate` to extract these metadata.

The actual data can be loaded with the `data` function. This will read the full
data which includes general information for each measurement along with the
metabolite concentrations and a quality information flag for each individual
measurement.

```{r}
metabo_data <- data(metabo)
dim(metabo_data)
```

In this data columns are variables (such as metabolite concentrations of quality
information) and rows participants. For the present data set concentrations for
175 metabolites are available along with quality information on each of these
measurements (hence there are in total 350 columns). The AIDs (i.e., identifiers
for each sample/participant) are provided as the row names of the
`data.frame`. Description and information on the individual columns can be
loaded with the `labels` function:

```{r}
metabo_ann <- labels(metabo)
```

This `data.frame` contains metadata and additional information on each variable
in `data`:

```{r}
head(metabo_ann)
```

Each row provides annotations for each column in the `metabo_data` `data.frame`
(rows and columns are in the same order).

```{r}
stopifnot(all(rownames(metabo_ann) == colnames(metabo_data)))
```

The column *cv_qc_chris* represent the coefficient of variation (CV) calculated
on QC samples, the QC CHRIS Pool samples. This values thus represents the
technical variability for each metabolite in the present dataset.

While these functions now loaded the data, it is suggested to further process
and reformat the data to simplify its analysis. At first we replace the internal
identifiers for CHRIS labels (i.e. starting with `x0pt*`) with more meaningful
column names.

```{r}
colnames(metabo_data) <- metabo_ann$description
rownames(metabo_ann) <- metabo_ann$description
```

With that it is much easier to access individual values.

```{r}
quantile(metabo_data$Gly, na.rm = TRUE)
```

It is also helpful to discriminate between columns in `metabo_data` that contain
the actual metabolite concentrations or the quality information. Below we
identify the columns with quality information. For these the keyword * flags* is
added to the metabolite name.

```{r}
flag_cols <- grep("flags$", colnames(metabo_data))
```

We next list all available quality information (which is encoded as a factor).

```{r}
levels(unlist(metabo_data[, flag_cols]))
```

Also, we will need at some point to *annotate* metabolites with some additional
annotations. For that we can use the `metabo_ann` data frame that represents the
annotation for the *labels* of the data module. Below we select 5 random
metabolites and extract their annotation from this data frame.

```{r}
ids <- sample(colnames(metabo_data)[-flag_cols], 5)
metabo_ann[ids, ]
```

We could also simply calculate the mean abundance of these 5 metabolites
across all available CHRIS participants using the code below:

```{r}
vapply(metabo_data[, ids], mean, numeric(1), na.rm = TRUE)
```

As we can see from the values above they are in **natural scale** - so, for data
analysis it might be better to transform them using `log2` or `log10` (which
will also ensure the data to be more Gaussian distributed). See also
[@verri_hernandes_age_2022] for more information on data distribution and
quality.


# Session information

```{r}
sessionInfo()
```

# References
