---
title: "Using the CHRIS Somalogic TDFF(ormat) data in R"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
---

```{r  biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
set.seed(123)
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Introduction

This document describes how the Somalogic SomaScan-based plasma proteomics data
from the CHRIS study can be loaded and analyzed in R. The data was reformatted,
from its original Somalogic *adat* file format into the *Textual Data File
Format* (TDFF) that is used as standard format for all CHRIS resources.

The Somalogic SomaScan data provides relative abundance of over 7000 proteins,
each analyte sometimes also referred to as a *SOMAmer*. The data was measured
using a microarray-based approach and has been fully normalized by Somalogic. No
apparent batch effect was observed (not for microarray slides, scanners used or
plates) on the present data.

The provided *values* for each SOMAmer (protein) is a *relative fluorescence
unit* (RFU) in **natural scale**.

The CHRIS somalogic data is available as two different modules:

- **chris_somalogic**: contains protein abundances for about 4,000 CHRIS
  participants. This data set does not contain any quality control or
  calibration sample. Samples not passing the Somalogic quality assessments were
  also removed. Also, all SOMAmer targeting non-human proteins have been
  removed. This data module should be used for regular data analyses.
- **chris_somalogic_full**: contains the *full* Somalogic data, i.e. also
  measurements in blank samples, calibration samples and QC samples (from
  Somalogic and pooled samples). No sample or SOMAmer was removed. This data
  set should only be used for quality assessment of the Somalogic data set.


# Usage

Data in TDF format is best imported into R using the
[tidyfr](https://github.com/EuracBiomedicalResearch/tidyfr) R package which can
be installed using the code below.

```{r, eval = FALSE}
remotes::install_github("EuracBiomedicalResearch/tidyfr")
```

To load the CHRIS Somalogic proteomics data access to the folder with the data
in TDF format is required. The name of the data folder (and hence the name of
the *CHRIS module*) is *chris_somalogic*. Below we list all versions of the data
set available for the this module. In this example we assume the
*chris_somalogic* folder to be present in the current folder in which R is
run. Parameter `path` allows to specify the folder in which CHRIS data modules
are stored (we use `"."` to indicate the current working directory).

```{r}
library(tidyfr)

list_data_modules(path = ".")
```

In the current folder thus one version for both Somalogic data modules are
available. We can next load the *chris_somalogic* module with the `data_module`
function specifying the name of the module, the version we want to load and also
the path where the module can be found.

```{r}
chris_soma <- data_module("chris_somalogic", version = "1.0.3.1", path = ".")
```

Loading the module can take some time because the function checks the complete
data set for its validity.  The `chris_soma` is now a reference to this module
which provides some general information:

```{r}
chris_soma
```

Note that we could also use the functions `moduleName`, `moduleVersion`,
`moduleDescription` and `moduleDate` to extract these metadata.

The actual data can be loaded with the `data` function. This will read the full
data which includes general information for each measurement along with the
quantified protein levels. Also, the `data` function will ensure that the
correct data types is assigned to the different variables. By default, the AIDs
(identifiers for the individual samples) are used as the row names of the
returned `data.frame`.

```{r}
soma_data <- data(chris_soma)
dim(soma_data)
```

In this data columns are variables (such as protein abundances) and rows
participants. Description and information on the individual columns can be
loaded with the `labels` function:

```{r}
soma_ann <- labels(chris_soma)
```

This `data.frame` contains metadata and additional information on each variable
in `data`:

```{r}
head(soma_ann)
```

As can be seen above, the first variables provide general information on the
measurement, such as the ID of the plate on which a sample was processed
(*PlateId*), the ID of the microarray scanner that was used to scan the slide
(*ScannerID*), the ID of the actual microarray (*SlideId*). As mentioned above,
no strong batch effects were seen on the data. Information on the actual
SOMAmers (i.e. annotation of the analyte to proteins) are provided in columns
*AptName* to *ColCheck*. Columns *cv_qc*, *cv_qc_chris*, *Dratio* and *eLOD*
represent the coefficient of variation (CV) calculated on QC samples, the CV
calculated on QC CHRIS Pool samples, the D-ratio calculated for each analyte and
the estimated level of detection (calculated using the method proposed in
[@candia_assessment_2022]. The Dratio represents the ratio between the
biological and technical variance [@broadhurstGuidelinesConsiderationsUse2018].

While these functions now loaded the data, it is suggested to further process
and reformat the data to simplify its analysis. At first we replace the internal
identifiers for CHRIS labels (i.e. starting with `x0so*`) with more meaningful
column names.

```{r}
colnames(soma_data) <- soma_ann[colnames(soma_data), "description"]
rownames(soma_ann) <- soma_ann$description
```

With that it is much easier to access individual values, such as the sample type
for each measured sample.

```{r}
table(soma_data$SampleType)
```

It is also helpful to identify the columns in `soma_data` that contain the
actual data on the SOMAmers (i.e. the relative protein abundances). Below we
identify these columns (which have a non-empty value in the *AptName* column of
the annotation `data.frame`).

```{r}
somamer <- soma_ann$AptName[soma_ann$AptName != ""]
```

Also, we will need at some point to *annotate* SOMAmers to the actual protein
identifiers. For that we can use the `soma_ann` data frame that represents the
annotation for the *labels* of the data module. Below we select 5 random
SOMAmers and extract their annotation from this data frame.

```{r}
ids <- sample(somamer, 5)
soma_ann[ids, ]
```

We thus get for these SOMAmers the protein they are designed to measure as well
as quality control data (coefficients of variation).

We could also simply calculate the mean abundance of these 5 SOMAmers across all
available CHRIS participants using the code below:

```{r}
vapply(soma_data[, ids], mean, numeric(1))
```

As we can see from the values above they are in **natural scale** - so, for data
analysis it might be better to transform them using `log2` or `log10` (which
will also ensure the data to be more Gaussian distributed).

At last we evaluate the signal distribution of the first selected SOMAmer both
in natural and log2 scale. In addition we indicate the estimated LOD for this
SOMAmer in the plot.

```{r}
id <- ids[1L]
par(mfrow = c(1, 2))
plot(density(soma_data[, id]), xlab = "abundance")
grid()
abline(v = soma_ann[id, "eLOD"])
plot(density(log2(soma_data[, id])), xlab = expression(log[2]~abundance),
     xlim = c(9, 16))
grid()
abline(v = log2(soma_ann[id, "eLOD"]))
```

For this SOMAmer all signals are thus above the estimated LOD.

At last we calculate for each SOMAmer the number of values below eLOD.

```{r}
nas <- vapply(somamer, function(z) {
    sum(soma_data[, z] < soma_ann[z, "eLOD"], na.rm = TRUE)
}, integer(1))
```

For in total `r sum(nas > 0)` of the `r length(nas)` at least one reported value
is below the eLOD of that SOMAmer. We can also calculate the percentage of
values below eLOD for each SOMAmer.

```{r}
perc_abs <- nas / sum(soma_data$SampleType == "Sample") * 100
```

In total `r sum(perc_abs > 50)` SOMAmers have more than 50% of the reported
values below their eLOD. Depending on the analysis performed, these might need
to be removed from the data analysis.


# Session information

```{r}
sessionInfo()
```

# References
