---
title: "Using the CHRIS ScanningSWATH TDFF(ormat) proteomics data in R"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
---

```{r  biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Introduction

This document describes how the ScanningSWATH (Mass Spectrometry)-based plasma
proteomics data from the CHRIS study can be loaded and analyzed in R. The data
was reformatted, from its original format provided by the proteomics facility of
the Charite in Berlin into the *Textual Data File Format* (TDFF) that is used as
standard format for all CHRIS resources.

The ScanningSWATH proteomics data provides (relative) abundances of about 150
plasma proteins quantified by a mass spectrometry (MS)-based approach. The data
set is already normalized and no apparent batch effects were observed. The data
is available at two different *levels*: the original abundances of the
individual peptides and the protein level data which represents the summarized
abundances of peptide intensities for each single protein.

The provided abundance values for proteins (or peptides) are provided in **log2
scale**.

The complete CHRIS ScanningSWATH data is available as four different modules:

- **chris_scanningswath_proteins**: contains protein abundances for about 4,000
  CHRIS participants. This data set does not contain any quality control
  samples. This data module should be used for regular data analyses.
- **chris_scanningswath_peptides**: contains peptide abundances for about 4,000
  CHRIS participants. This data set does not contain any quality control
  samples.
- **chris_scanningswath_proteins_full**: contains the *full* protein-level data,
  i.e. also measurements in QC samples. This data set should only be used for
  quality assessment purposes.
- **chris_scanningswath_peptides_full**: contains the *full* peptide-level data,
  i.e. also measurements in QC samples. This data set should only be used for
  quality assessment purposes.



# Usage

Data in TDF format is best imported into R using the
[tidyfr](https://github.com/EuracBiomedicalResearch/tidyfr) R package which can
be installed using the code below.

```{r, eval = FALSE}
remotes::install_github("EuracBiomedicalResearch/tidyfr")
```

To load the CHRIS ScanningSWATH proteomics data, access to the folder with the
data in TDF format is required. The name of the data folder (and hence the name
of the *CHRIS module*) is *chris_scanningswath_proteins*. Below we list all
versions of the data set available for the this module. In this example we
assume the *chris_scanningswath_proteins* folder to be present in the current
folder in which R is run. Parameter `path` allows to specify the folder in which
CHRIS data modules are stored (we use `"."` to indicate the current working
directory).

```{r}
library(tidyfr)

list_data_modules(path = ".")
```

In the current folder we have thus one version for all 4 ScanningSWATH
modules. We next load the *chris_scanningswath_proteins* module (which should be
used for *standard* analyses) with the `data_module` function specifying the
name of the module, the version we want to load and also the path where the
module can be found.

```{r}
chris_prot <- data_module("chris_scanningswath_proteins",
                          version = "1.0.0.1", path = ".")
```

The `chris_prot` is now a reference to this module which provides some general
information:

```{r}
chris_prot
```

Note that we could also use the functions `moduleName`, `moduleVersion`,
`moduleDescription` and `moduleDate` to extract these metadata.

The actual data can be loaded with the `data` function. This will read the full
data which includes general information for each measurement along with the
quantified protein levels.

```{r}
prot_data <- data(chris_prot)
dim(prot_data)
```

In this data columns are variables (such as protein abundances) and rows
participants. Description and information on the individual columns can be
loaded with the `labels` function:

```{r}
prot_ann <- labels(chris_prot)
```

This `data.frame` contains metadata and additional information on each variable
(column) in `data`. Below we list the last 6 rows of that `data.frame`.

```{r}
tail(prot_ann)
```

Thus, in addition to quantified protein levels we have also general *metadata*
variables providing information on the instrument on which the data was
generated or the originating plate for the samples. Variables representing
protein measurements will have non-empty information in column
`"Protein.Ids"`. The last variable `"cv_qc_chris"` provides in addition the
coefficient of variation for each protein calculated across (repeatedly
measured) QC CHRIS Pool samples and hence provides information on the technical
variance of measurements for each protein.

While the `data` and `labels` functions allow to load the data, it is suggested
to further process and reformat it to simplify its analysis. At first we replace
the internal identifiers for CHRIS labels (i.e. starting with `x0sw*`) with more
meaningful column names.

```{r}
rownames(prot_ann) <- prot_ann$description
colnames(prot_data) <- prot_ann$description
```

With that it is much easier to access individual values.

```{r}
table(prot_data$MS_Batch)
```

It is also helpful to identify the columns in `prot_data` that contain the
actual data on the proteins (i.e. the summarized protein abundances). Below we
identify these columns (which have a non-empty value in the *Protein.Ids* column
of the annotation `data.frame`).

```{r}
prot_cols <- rownames(prot_ann[prot_ann$Protein.Ids != "", ])
```

In the data set we have thus protein abundances for in total
`r length(prot_cols)` proteins. The `prot_cols` variable can now also be helpful
to extract the actual protein values. Below we for extract for example the CV
for all proteins and plot their distribution.

```{r}
prot_cvs <- prot_ann[prot_cols, "cv_qc_chris"]
plot(density(prot_cvs), xlab = "CV")
grid()
abline(v = 0.3)
```

Thus, most of the proteins have a CV lower than 0.3 (30% technical variance;
vertical black line in the plot above).

Below we calculate for each protein the mean value across all samples in the
data set to show how the `prot_cols` information can be used to subset the
`prot_data` `data.frame`.

```{r}
mean_conc <- vapply(prot_data[, prot_cols], mean, numeric(1))
plot(sort(mean_conc), ylab = expression(log[2]~abundance))
grid()
```

As we can see from the plot above the values are in **log2 scale**.


# Session information

```{r}
sessionInfo()
```

# References
