# Export the Biocrates Metabolomics data

Here we export the Biocrates-based targeted metabolomics data from CHRIS (and
NAFLD) in the new Textual Dataset Format (TDF) using the `tidyfr` package.

```{r}
library(tidyfr)
library(biochristes7500)

data("biochristes7500")
```

Extract the *data*, *labels* and *groups* from the `SummarizedExperiment`.

```{r}
data <- data(biochristes7500, assayNames = c("concentrations", "flags"),
             labelPrefix = "x0pt")

data("qcs_fia")
data("qcs_lcms")

qc_types <- unique(qcs_fia$sample_name)
cv_fia <- lapply("QC CHRIS Pool", function(z) {
    rowRsd(concentrations(qcs_fia[, qcs_fia$sample_name == z],
                          blessing = "none"))
})
cv_fia <- unlist(cv_fia)

cv_lcms <- lapply("QC CHRIS Pool", function(z) {
    rowRsd(concentrations(qcs_lcms[, qcs_lcms$sample_name == z],
                          blessing = "none"))
})
cv_lcms <- unlist(cv_lcms)

cvs <- rep(NA_real_, nrow(biochristes7500))
cvs[match(names(cv_fia), rownames(rowData(biochristes7500)))] <- cv_fia
cvs[match(names(cv_lcms), rownames(rowData(biochristes7500)))] <- cv_lcms

rowData(biochristes7500)$cv_qc_chris <- cvs


labels <- labels(
    biochristes7500, assayNames = c("concentrations", "flags"),
    labelPrefix = "x0pt")
labels$long_description <- labels$biochemical_name
labels$long_description[grep("flags$", labels$description)] <-
    paste0(labels$long_description[grep("flags$", labels$description)],
           " flags")

groups <- groups(
    biochristes7500, assayNames = c("concentrations", "flags"),
    labelPrefix = "x0pt")
```

We next replace the integer-type flags with `factor`s to better encode the
quality information.

```{r}
cats <- c(`0` = "OK", `-9` = "Removed because of technical reason",
          `-1` = "Below lower level of quantification",
          `1` = "Above upper level of quantification",
          `-2` = "Below level of detection")
idx <- grep("a$", colnames(data))
for (i in idx)
    data[, i] <- unname(factor(cats[as.character(data[, i])], levels = cats))
```

Fixing and adapting the *labels*.

```{r}
labels$type[labels$type == "integer"] <- "categorical"
labels <- labels[, !colnames(labels) %in%
                   c("analyte_lloq", "analyte_uloq", "analyte_lod",
                     "analyte_blessed_min", "analyte_blessed_max")]
```

Adding a grouping for each concentration-quality pair.

```{r}
grp_labels <- data.frame(group = unique(groups$group),
                         description = unique(groups$group))
grp_labels$description <- sub("analyte_", "Values for ",
                              grp_labels$description)
grp_labels$description[1L] <- "Metabolite concentrations"
grp_labels$description[2L] <- "Measurement quality information"
```

Exporting the data.

```{r}
meta <- metadata(biochristes7500)
data_version <- meta$value[meta$name == "version"]
data_date <- meta$value[meta$name == "data_date"]

export_tdf(name = "metabolomics_p180",
           description = "Targeted metabolomics data based on the Biocrates p180 kit comprising measured concentrations of 175 metabolites and lipids in serum samples of CHRIS participants.",
           version = "1.0.1.0",
           date = data_date,
           path = "/home/jo/data/CHRIS/chris_baseline",
           data = data, groups = groups, grp_labels = grp_labels,
           labels = labels)
```

## Evaluating export

Check that data is exported correctly.

```{r}
bp <- "/home/jo/data/CHRIS/chris_baseline"

list_data_modules(bp)

mdl <- data_module(name = "metabolomics_p180", version = "1.0.1.0", path = bp)
mdl

head(labels(mdl))
head(groups(mdl))
head(grp_labels(mdl))
head(data)

tmp <- data(mdl)

mat <- as.matrix(t(tmp[, seq_len(nrow(biochristes7500)) + 1]))
library(testthat)
expect_equal(unname(mat), unname(assay(biochristes7500)))

flags <- as.matrix(t(tmp[, seq_len(nrow(biochristes7500)) + 176]))
flags[flags == "OK"] <- 0
flags[flags == "Removed because of technical reason"] <- -9
flags[flags == "Below lower level of quantification"] <- -1
flags[flags == "Above upper level of quantification"] <- 1
flags[flags == "Below level of detection"] <- -2
flags <- apply(flags, c(1, 2), as.numeric)

expect_equal(unname(flags), unname(assay(biochristes7500, "flags")))
```

# Session information

```{r}
sessionInfo()
```
