---
title: "Import and reformatting of CHRIS Somalogic SomaScan data"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
---

```{r  biocstyle, echo = FALSE, results = "asis" }
library(BiocStyle)
BiocStyle::markdown()
```

**Compiled**: `r date()`

# Import and processing of Somalogic SomaScan data

## Background

Somalogic assays base on protein-capture Slow Offrate Modified Aptamer (SOMAmer)
reagents [@candia_assessment_2017]. SOMAmer reagents are designed against a
variety of human proteins. For serum and plasma samples, SOMAmer reagents are
binned into 3 separate groups according to the expected endogenous abundance of
each cognate protein (each SOMAmer thus exists only in one of the 3
dilutions). Serum and plasma samples (including controls) are then diluted into
3 concentrations. The diluted samples are then incubated with the
dilution-specific SOMAmers. A typical SOMAscan plate will include buffer wells
(no sample added), quality control (QC_SOMAscan) and calibrator samples specific
for the used sample matrix (plasma or serum). The calibrator sample (5-7
replicates per plate) consists of a common pooled sample used across all plates
with the purpose to reduce between plate differences.


## Normalization

Raw data obtained from slide-based hybridization microarrays exhibit intraplate
variances (differences in loading volume, leaks, washing conditions) as well as
batch effects across plates. Somalogic defines different normalization steps
addressing all these variances. Descriptions for this methods are extracted from
the Somalogic technical documentation and from [@candia_assessment_2017].

- Hybridization Control Normalization (*HybNorm*): remove systematic
  (hybridization) biases of individual wells. It is performed using a set of 12
  hybridization control elution (HCE) spike-ins that are added to each
  sample. The measured Relative Fluorescence Intensities/Units (RFU)s of these
  are compared against a reference which is determined by the median of such
  spike-ins across the plate [@candia_assessment_2017]. Each sample is
  normalized with a single scale factor (median ratio between the observed and
  the reference RFU values of the HCE).

- Intraplate Median Signal Normalization (*medNormInt*): aims to remove any
  variances related to overall protein concentration, pipetting variation,
  differences in concentrations and other variances within a single
  plate. Normalization is performed separately for each dilution set and wells
  of the same sample class (i.e. QC type, calibrator, biological sample). A
  detailed description of the method is given in
  [@candia_assessment_2017]. According to the *Somalogic SomaScan Data Standard
  Tech Note* this is only performed on calibrator samples.

- Plate scaling and calibration (*plateScale*): between plate (array)
  differences are adjusted using a number of replicate measurements of a common
  pooled calibrator sample which has to be the same across plates and which has
  to be consistent with the assay sample type of a study. A unique scale factor
  is defined for each SOMAmer binding reagent. The median of these is then
  computed (within the plate?) and applied to all measurements in that plate
  (plate scaling). The scale factors are subsequently recalculated for each
  SOMAmer and applied to all measurements within the set of samples in that
  plate.

- Median signal normalization to a reference: is performed for each sample. The
  scale factor for a set of SOMAmer reagents is computed against a reference
  value generated from a cohort of healthy normal individuals and then
  aggregated within a dilution. The median of each dilution's scale factors is
  then applied to the respective SOMAmer reagents.

## R packages

There was a `readat` Bioconductor package that however is no longer available on
Bioconductor. It however can be installed from github (*graumannlab/readat*).

Alternatively, there is the `SomaDataIO` package (github:
*SomaLogic/SomaDataIO*) that can be used for data import.


## Data import

We next read the data to evaluate how the data is represented.

```{r}
library(SomaDataIO)
```

Different files (assuming for different stages of normalization) are available:

```{r}
path <- "/home/jo/data/CHRIS/proteomics-somalogic/"
fls <- dir(path, pattern = "adat", full.names = TRUE)
basename(fls)
```

We load the most processed data file (`r basename(fls)[8]`) since that is also
the one suggested to be used according to the [Somalogica Quality
Statement](SS-2231571_SQD.pdf).

```{r}
dta <- read_adat(fls[8L])
```

Feature (i.e. SOMAmer) annotations (column data) can be extracted with
`getAnalyteInfo`.

```{r}
anlts <- getAnalyteInfo(dta) |> as.data.frame()
head(anlts)
nrow(anlts)
```

This information includes also normalization factors and quality assessments
performed for individual SOMAmers. Maybe interesting might be the `"ColCheck"`
column that provides information on the quality of the specific SOMAmer:

```{r}
table(anlts[, "ColCheck"])
```

There are quite some features that did not pass the QC. According to Somalogic
*Flagged SOMAmer reagents are typically retained for analyses since accuracy
across all assay runs is a robust quality metric but is not a requirement for
identification of meaningful biological signal.* Thus, we will keep all these
features but they might be treated with care in the final interpretation of
results.

Note that there are different `"Type"` of SOMAmers that are also measured in
different dilutions.

```{r}
table(anlts[, "Type"])
```

We might want to exclude the `"Hybridization Control Elution"` features from the
*final* data set for actual data analysis (but keep them in the full data set
for QC).

The rest seems to be one huge `data.frame`/`tibble`: each row being one sample,
with columns being annotations, metadata information or aptamer.

```{r}
rfu <- dta[, anlts$AptName]
all(vapply(rfu, is.numeric, logical(1)))
```

Check the distribution of mean RFU per SOMAmer.

```{r}
boxplot(log2(vapply(rfu, mean, numeric(1))),
        ylab = expression(log[2]~RFU))
```

Checking the available annotation for samples.

```{r}
row_ann <- dta[, !(colnames(dta) %in% anlts$AptName)] |> as.data.frame()
head(row_ann)
```

The `"RowCheck"` column provides information whether the row (sample) passed the
quality criteria:

```{r}
table(row_ann$RowCheck)
```

These samples should be excluded from statistical data analysis. We will remove
them from the *final* data set but keep all samples for the full data set for
quality assessment.


## Map IDs to AIDs

We need to map the provided IDs from Somalogic to the AIDs from CHRIS.

```{r}
map <- read.table(file.path(path, "mapping.txt"), sep = "\t", header = TRUE,
                  colClasses = c("character", "character"))
idx <- match(map$sample_ID, row_ann$Barcode2d)

row_ann$aid <- NA_character_
row_ann$aid[idx] <- map[, "AID"]
row_ann$SampleType[which(row_ann$aid == "POOL")] <- "QC_CHRIS"

stopifnot(all(row_ann$Barcode2d[idx] == map[, "sample_ID"]))
```

The table below lists the total numbers for the different sample types (before
excluding flagged samples).

```{r}
table(row_ann[, "SampleType"])
```


# Calculate quality metrics for measurements

We next calculate coefficients of variation (CV) and levels of detection (LOD)
for each SOMAmer in the data set. These are calculated on the subset of samples
that passed the Somalogic quality assessment.

```{r}
data_qc <- as.data.frame(dta)
stopifnot(all.equal(rownames(data_qc), rownames(row_ann)))
data_qc$aid <- row_ann$aid
data_qc$SampleType <- row_ann$SampleType
data_qc <- data_qc[data_qc$RowCheck == "PASS", ]
```

At first we calculate the CV using Somalogic or CHRIS Pool QC samples.

```{r}
rsd <- function(x, na.rm = TRUE) {
    sd(x, na.rm = na.rm) / abs(mean(x, na.rm = na.rm))
}
## QC
cv_qc <- vapply(data_qc[data_qc$SampleType == "QC", anlts$AptName],
                rsd, numeric(1))
cv_qc_chris <- vapply(data_qc[data_qc$SampleType == "QC_CHRIS", anlts$AptName],
                      rsd, numeric(1))
```

In addition we calculate the dispersion ratio (D-ratio) that represents the
ratio between the biological and technical variance
[@broadhurstGuidelinesConsiderationsUse2018]. For an estimate of the technical
variance we use the standard deviation in QC CHRIS Pool samples and the data is
first log transformed as suggested in
[@broadhurstGuidelinesConsiderationsUse2018].

```{r}
fun <- function(z) {
    sd(log2(z), na.rm = TRUE)
}
dratio <-
    vapply(data_qc[data_qc$SampleType == "QC_CHRIS", anlts$AptName],
           fun, numeric(1)) /
    vapply(data_qc[data_qc$SampleType == "Sample", anlts$AptName],
           fun, numeric(1))
```

We next calculate also the level of detection (LOD) for each SOMAmer using the
method proposed by Candia et al. [@candia_assessment_2022].

```{r}
elod <- function(x) {
    median(x, na.rm = TRUE) + 5 * mad(x, na.rm = TRUE)
}
elods <- vapply(data_qc[data_qc$SampleType == "Buffer", anlts$AptName],
                elod, numeric(1))
```

At last we add these quality information to the data.

```{r}
anlts$cv_qc <- cv_qc
anlts$cv_qc_chris <- cv_qc_chris
anlts$Dratio <- dratio
anlts$eLOD <- elods
```


# Export data in TDFF(ormat)

The CHRIS Somalogic data set will be exported as two different *modules*, one
with only the (good quality) data for CHRIS participants and one including also
all calibrator features and all QC and calibrator samples. The latter will be
used for quality assessment, the former for actual data analyses.

```{r}
library(tidyfr)
```

First we create the *full* data set with all available data. We need also to
encode the variables with the coding scheme used in CHRIS.

```{r}
data <- as.data.frame(dta)
stopifnot(all.equal(rownames(data), rownames(row_ann)))
## Convert some variables to factors.
data$PlateId <- factor(data$PlateId)
data$ScannerID <- factor(data$ScannerID)
data$SlideId <- factor(data$SlideId)
data$SampleType <- row_ann$SampleType
data_cn <- colnames(data)
sid <- data$SampleId

colnames(data) <- sprintf("x0so%04d", seq_len(ncol(data)))
## Add AIDs
data$aid <- row_ann$aid
nas <- is.na(data$aid)
data$aid[nas] <- sid[nas]
data <- data[, union("aid", colnames(data))]
```

Next we define the *labels*.

```{r}
lbls <- labels_from_data(data)
lbls$description <- data_cn
lbls$long_description <- data_cn
```

Add annotation to the individual labels (for SOMAmers). We only add selected
annotations and not all calibration values etc to the data.

```{r}
ann_cn <- c("AptName", "SeqId", "SeqIdVersion", "SomaId", "TargetFullName",
            "Target", "UniProt", "EntrezGeneID", "EntrezGeneSymbol",
            "Organism", "Units", "Type", "Dilution", "ColCheck",
            "cv_qc", "cv_qc_chris", "Dratio", "eLOD")

rownames(anlts) <- anlts$AptName
lbls <- cbind(lbls, anlts[lbls$description, ann_cn])
lbls$long_description <- anlts[lbls$description, "TargetFullName"]
lbls[is.na(lbls$long_description), "long_description"] <-
    lbls$description[is.na(lbls$long_description)]
```

Next we define *groups*, i.e. variables that can be grouped into the same
entity.

```{r}
grps <- data.frame(group = "metadata", label = lbls$label)
grps[grep("^seq", lbls$description), "group"] <- "protein_abundance"

grplbls <- data.frame(group = c("metadata", "protein_abundance"),
                      description = c("Sample and plate information",
                                      "Protein abundance (expressed as RFU)"))

```

Next we export the full data set.

```{r}
export_tdf(
    name = "chris_somalogic_full",
    description = paste0("CHRIS Somalogic SomaScan-based plasma proteomics ",
                         "data; full data including quality controls and ",
                         "calibrator samples"),
    version = "1.0.3.0", date = "2023-03-03",
    data = data, groups = grps, grp_labels = grplbls, labels = lbls)
```

Next we subset the data to CHRIS study samples (excluding QC CHRIS Pool) and
samples that were passing the Somalogic quality control. In addition we remove
all non-human proteins from the data set.

```{r}
rem <- anlts$AptName[!anlts$Organism %in% c("Human")]
lbls <- lbls[!lbls$description %in% rem, ]
data <- data[, c("aid", lbls$label)]
grps <- grps[grps$label %in% lbls$label, ]

## Remove QC control samples and QC CHRIS Pool samples.
data <- data[grep("^001", data$aid), ]

## Remove samples not passing QC.
data <- data[data$x0so0027 == "PASS", ]
```

And finally exporting the data.

```{r}
export_tdf(
    name = "chris_somalogic",
    description = paste0("CHRIS Somalogic SomaScan-based plasma proteomics ",
                         "data; CHRIS study samples passing Somalogic QC."),
    version = "1.0.3.0", date = "2023-03-03",
    data = data, groups = grps, grp_labels = grplbls, labels = lbls)
```


# Session information

```{r}
sessionInfo()
```

# References
